---
editor_options: 
  markdown: 
    wrap: 72
---

**University of Edinburgh**

**School of Mathematics**

**Bayesian Data Analysis, 2023/2024, Semester 2**

**Workshop 3: Introduction to Stan**

**Note**: Before starting this practical, you are advised to spend some
time looking at the Stan examples we have discussed during Lecture 4
(code is available on Learn). If you already looked at these, then feel
free to go directly to question 1. The code below loads Stan.

```{r}
library(rstan)
options(mc.cores = parallel::detectCores())

#Installing Stan can be done by the next line
#install.packages("rstan", repos = c("https://mc-stan.org/r-packages/", getOption("repos")))

#If you have encountered some issues do to previous attempts to install Stan,
#run the next line first to remove previous versions of rstan 
#remove.packages(c("StanHeaders", "rstan"))

#Test whether your installation works
#example(stan_model, package = "rstan", run.dontrun = TRUE)
```

**1. Analysis of binomial data: revisiting the drug example**

**A new drug is being considered for relief of chronic pain, with the
success rate** $\theta$ **being the proportion of patients experiencing
pain relief (i.e. number of successes has binomial distribution).
According to past information, a Beta(9.2, 13.8) prior distribution was
suggested for** $\theta$. This drug had 15 successes out of 20 patients.

**1.1. Compute the posterior mean, standard deviation and a** $95\%$
**credible interval.**

```{r}

```

**1.2. What is the probability that the true success rate is greater
than 0.6.**

```{r}

```

**1.3. Suppose 40 more patients were entered into the study. What is the
chance that at least 25 of them experience pain relief?**

```{r}

```

**2. State space modeling for gray whale abundance**

**In this exercise, we continue work with our state space model from
Lecture 4.**

**As a recap, we have some observations of gray whale abundance (i.e.
population size) during some of the years in the period 1951-1997. It
turns out that it is easier to fit a model on the logarithm of the
population rather than the population directly. We are going to use the
following models from the true log-population** $x_t$ **(**$t$ **denotes
the year),**

$$x_t = b x_{t-1}+u+w_t; \quad w_t\sim N(0,\sigma^2).$$

**The observations are not assumed to be exact, but we allow them to
have an error, this is modelled as**

$$y_t =x_{t}+v_t, \quad v_t\sim N(0,\eta^2).$$

**This is a simple State-Space Model (SSM), also called Hidden Markov
Model.** $x_t$ **are the hidden states, and** $y_t$ **are the
observations.**

**The following lines load the dataset and create the observation
vector** $y$.

```{r}
library("MARSS")
#This loads to MARSS library
#If not available, please use install.packages("MARSS") first before loading it

# We load the graywhales dataset, and print out the values
data(graywhales)
print(graywhales)

n=1997-1951+1;
#We will have one y[i] for each year from 1951 until 1997
#So i=1 corresponds to 1951, i=2 corresponds to 1952, etc.

y=rep(NA,n);
for(i in 1:nrow(graywhales)){
    y[graywhales[i,1]-1950]=log(graywhales[i,2]);
}
```

**2.1. Implement the model in Stan using the following prior
distributions for the model parameters:**

-   **Gaussian prior** $N(\log(2500),1)$ for the initial state $x_0$,

-   **Uniform prior on the interval (0,1) for parameter** $b$,

-   **Exponential prior with parameter 1 for** $u$,

-   **Inverse Gamma prior with parameters (0.1,0.1) for** $\sigma^2$ and
    $\eta^2$.

**Hint: you can use the code provided with Lecture 4 for this.**

**Compile and run the Stan simulations. Report the effective samples
sizes for these 5 parameters. Choose the number of steps in the burn-in
period and the number of MCMC iterations in a way to ensure that the
effective sample sizes for the 4 parameters** $x_0$,$b$**,** $u$**,**
$\sigma^2$ **and** $\eta^2$ **are all above 1000. Once this is ensured,
compute summary statistics and plot the posterior densities of these
parameters.**

```{r}

```

**2.2. Now we are going to perform prior sensitivity checks. Try a
different prior distribution for the 5 variables. For** $x_0$**, please
do not change the mean significantly from log(2500), but you can try a
different variance. When formulating priors, you are encouraged to use
your understanding of the model parameters and biological facts (see
e.g. <https://en.wikipedia.org/wiki/Gray_whale>). Run the simulations
again, and compare the summary statistics and posterior density plots of
these 5 parameters.**

```{r}

```

**2.3.** **Update the model to compute the posterior distribution of the
log population sizes (**$x_t$**) every year between 1951-2050. Plot the
evolution of the posterior mean of the log population sizes from
1951-2050 on a single plot, along with two other curves that correspond
to the [2.5%, 97.5%] credible interval of the log population size
(**$x_t$**) according to the posterior distribution at each year.
Finally, estimate the posterior probability that the population of gray
whales becomes smaller than 100 at any year from 1951 until the end of
2050 (i.e.,** $p(\min_{t\in \{0,1\ldots,99\}} x_t<=\log(100)|y)$**).**

```{r}

```

**2.4. We are going to perform posterior predictive checks to evaluate
the fit of this model on the data. First, create replicate observations
from the posterior predictive using Stan. The number of replicate
observations should be at least 1000. Compute the minimum, maximum,
median, kurtosis and skewness for the replicated observations. Plot the
histograms for these quantities together with a line that shows the
value of the function considered on the actual dataset (see the R code
for Lecture 3 for an example). Perform these checks both for the
original prior from Question 2.1, and the prior you have proposed in
Question 2.2. Discuss the results.**

```{r}

```

**3. Multivariate Multiple Regression - Seemingly Unrelated
Regressions**

**In this exercise, we are going to work on the high school exam scores
dataset from Lecture 3.** **We start by loading the dataset, and fitting
regression models on all exam scores ()**

```{r}
#Loading the dataset
df<-read.csv("sur_scores.csv",header = TRUE, sep = ",")
head(df)
```

```{r}
df$prog=as.factor(df$prog)
#converting prog it to factor format (categorical variable)
#the rest of the covariates are textual, so automatically treated as categorical
m1<-lm(cbind(math,read, write, science, socst)~female+race+ses+schtyp+prog,data=df)
#socst refers to social studies
#when using the cbind(math,read, write, science, socst) as response, two independent linear models are fitted for each response
summary(m1)
```

**3.1. Fit a Bayesian Multivariate Multiple Regression model in Stan for
these response variables using female, race, ses, schtyp, and prog as
covariates (see the code of Lecture 4, available on Learn).**

**Assume that the covariance matrix between the error terms is diagonal
(i.e. the regressions are independent).**

**Print out the summary statistics, and compute the RMSE values over the
whole dataset.**

**3.2. Set a prior distribution on the error covariance matrix that
allows for non-diagonal matrices too (one possible way is by setting a
prior on the correlation matrix and the eigenvalues of the covariance
matrix separately, see the code of Lecture 4). Re-fit your Bayesian
Multivariate Multiple Regression model using this prior, i.e. allow for
non-diagonal covariances.**

**Print out the summary statistics, and compute the RMSE values over the
whole dataset.**

**3.3. Perform posterior predictive checks for both models. Discuss the
results.**

**4. Infectious disease model.**

**In this excercise, we are going to implement a Bayesian infectious
disease model, based on the paper "Bayesian workflow for disease
transmission modeling in Stan'' by Grinsztajn et al.**

We will use a simple SIR model (Susceptible/Infected/Recovered), which
is a population based model (i.e. individual characteristics are not
taken into account).

Let $N$ denote the total size of the population (assumed to stay
constant).

Let $S$ , $I$, and $R$ denote the number of Susceptible (i.e. they can
be infected), Infected, and Recovered persons. These depend on the time
parameter $t$.

The evolution of these is modeled with the following ODE (ordinary
differential equation):

$$
\begin{aligned}
 \frac{dS}{dt} &= -\beta  S \frac{I}{N}\\
 \frac{dI}{dt} &= \beta  S  \frac{I}{N} - \gamma  I \\
 \frac{dR}{dt} &= \gamma I
\end{aligned}
$$ where

-   $S(t)$ is the number of people susceptible to becoming infected (no
    immunity),

-   $I(t)$ is the number of people currently infected (and infectious),

-   $R(t)$ is the number of recovered people (we assume they remain
    immune indefinitely),

-   $\beta$ is the constant rate of infectious contact between people,

-   $\gamma$ the constant recovery rate of infected individuals.

There are certain assumptions required for this model to be a reasonable
approximation of infection spread (such as no births or deaths, i.e.
$S(t)+I(t)+R(t)=N$ for all times), see the original paper for more
details.

In general, infections start from one individual, we choose the initial
conditions $$I(0)=1, \quad R(0)=0, \quad S(0)=N-1.$$

In addition to the mathematical model itself, we also allow for
observation noise, by modelling the observed infections as

$$ I_\text{obs}(t) \sim \text{NegBin}(I_\text{ODE}(t),
\phi).$$

The Negative Binomial distribution has two parameters, it's mean
($I_\text{ODE}(t)$) and overdispersion parameter ($\phi$), see

<https://mc-stan.org/docs/functions-reference/nbalt.html>

We are going apply this model to a small dataset. This is about
influenza infections among students in a boarding school in England in
1978, from 1978-01-22 until 1978-02-04. There were $N=763$ students in
the school in total, 512 of them got infected at some point during this
period. The dataset is available as part of the outbreaks package.

```{r}
library(outbreaks)
print.data.frame(influenza_england_1978_school)
```

**4.1.** **Implement the above model on this dataset by assuming that
\$I(0)=1\$, and the observations** $I_\text{obs}(t)$ **for**
$t=1,\ldots,14$ **correspond to the in_bed column in the dataset. You
can either choose priors for the model parameters yourself, or choose
your priors as suggested in the original paper "Bayesian workflow for
disease transmission modeling in Stan''.**

**Hint: Since this process is defined in continuous time, and the
observations are in discrete time, you will need to solve the ODE
governing the dynamics within your Stan model, and then generate the
observations based on the solution of this ODE.**

**To do this, you will first need to define a function that computes the
derivatives (i.e. right hand side of the ODE). Such functions need to go
to the function{} block, that comes before any other blocks (such as
data, etc.), a simple example:**

```         
functions {

real add_up(real a, real b){
  return a + b;
}

}
```

**You can see how to create user defined functions here:
<https://mc-stan.org/docs/reference-manual/function-definition-block.html>**

**The ODE itself can be solved using the ODE_rk45 solver, which has the
following parametrization,**

`array[] vector` `ode_rk45(function ode, vector initial_state, real initial_time, array[] real times, ...)`**\
See
<https://mc-stan.org/docs/functions-reference/functions-ode-solver.html>
for more options.**

**Here ode is the function specifying the right hand side of the ODE.
This has to be in a specific format:**

```         
vector ode(real t, vector y, ...);
```

**The first argument to the system function is time, passed as a `real`;
the second argument to the system function is the system state, passed
as a `vector`, and the return value from the system function are the
current time derivatives of the state defined as a `vector`. Additional
arguments can be included in the system function to pass other
information into the solve (these will be passed through the function
that starts the ODE integration). These argument can be parameters,
data, or any quantities that are needed to define the differential
equation.**

**For some examples, see
<https://mc-stan.org/docs/stan-users-guide/coding-the-ode-system-function.html>.**

**4.2. Compile the model, run the MCMC sampler, and print out the
summary statistics from the model parameters. Plot the modeled number of
infections per day** $I(t)$ **against the observed number of
infections** $I_{obs}(t)$.

**4.3. Evaluate the predictive performance of this model by repeatedly
fitting it on the data from days** $1,\ldots, t$ **and predict the
number of infections on days** $t+1$, from $t=3$ **until** $t=13$**.
Plot the estimated infections versus the observed values.**

**4.4. Perform posterior predictive model checks. Discuss your
results.**
