---
editor_options:
  markdown:
    wrap: 72
output: pdf_document
---

**University of Edinburgh**\
**School of Mathematics**\
**Bayesian Data Analysis, 2023/2024, Semester 2**\

**Workshop 5: Temporal and Spatial Random Effects**

```{r}
library(INLA)
```

# 1. **Modelling fatal airline accidents from 1976 through 2001.**

**This exercise has been taken largely from a shortcourse at the
University of Copenhagen which occurred in January 2013 and notes from
Gurrin, Carstensen, Hojsgaard, and Ekstrom. The dataset `airline.RData`
is available on Learn.**

**The fields are:**\
**- Year1975 (number of years after 1975),**

**- Year,**

**- Fatal (number of fatal airline accidents),**

**- Miles (total passenger miles, in** $10^{11}$ **miles, e.g.,**
$3.863 = 3.683*10^{11} \text{miles} = 368.3$ **Billion miles),**

**- Rate (fatalities per** $10^{11}$ **passenger miles).**

**You will be fitting some temporal random effect Poisson models to
Fatal.**

```{r}
load("airline.RData")
head(airlines)
```

**1.1. Fit a simple Poisson model on Fatal in INLA with Year as a
covariate, with offset set as log(miles) (see Workshop 4, question 1.5),
i.e. a Poisson model of the form**
$$\log(\mu(t))=\beta_0 + \beta_1 (t-\bar{t})+\log(\text{miles}(t))$$
**Here miles is the total number of miles of flights per year divided by
10\^11 (ranging from 5 to 20 during the period 1976-2001.**

**Either choose your own priors for** $\beta_0$ and $\beta_1$, or use
the default priors.

**Compute the DIC, NLSCPO, and WAIC scores (WAIC can be computed by
setting waic=T in control.compute).**

```{r}
prior.beta=list(mean.intercept = 0, prec.intercept = 0.1,
                    mean = 0, prec = 1)
q1.m1.inla <- inla(fatal ~ 1+scale(year1975,center = TRUE, scale = FALSE),family="Poisson",data=airlines,offset=log(miles), control.fixed=prior.beta,control.compute=list(cpo=T,dic=T,waic=T))
summary(q1.m1.inla)
```

```{r}
cat("Marginal log-likelihood of model 1:",q1.m1.inla$mlik[1],"\n")
cat("DIC of model 1:",q1.m1.inla$dic$dic,"\n")
cat("WAIC of model 1:",q1.m1.inla$waic$waic,"\n")
cat("NSLCPO of model 1:",-sum(log(q1.m1.inla$cpo$cpo)),"\n")
```

```{r}
plot(airlines$year,q1.m1.inla$summary.fitted.values$sd,type="l",xlab="Year",ylab="sd(mu(t))")
```

```{r}
plot(airlines$year,q1.m1.inla$summary.fitted.values$mean,type="l",xlab="Year",ylab="E(mu(t))")
```

**1.2. Fit a Poisson model on Fatal in INLA with Year as a covariate,
with offset set as log(miles) (see Workshop 4, question 1.5), using RW1
random effect on the year, i.e. a Poisson model of the form**
$$\log(\mu(t))=\beta_0 + \beta_1 (t-\bar{t})+\log(\text{miles}(t))+f(t).$$**Compute
the DIC, NLSCPO, and WAIC scores. Plot the evolution of the posterior
standard deviation of the fitted values (**$\mu(t)$**) over the years
1976-2001.**

```{r}
prior.beta=list(mean.intercept = 0, prec.intercept = 0.1,
                    mean = 0, prec = 1)
q1.m2.inla <- inla(fatal ~ 1+scale(year1975,center = TRUE, scale = FALSE)+f(year1975,model="rw1",constr=TRUE),family="Poisson",data=airlines,offset=log(miles), control.fixed=prior.beta,control.compute=list(cpo=T,dic=T,waic=T))
summary(q1.m2.inla)
```

```{r}
cat("Marginal log-likelihood of model 2:",q1.m2.inla$mlik[1],"\n")
cat("DIC of model 2:",q1.m2.inla$dic$dic,"\n")
cat("WAIC of model 2:",q1.m2.inla$waic$waic,"\n")
cat("NSLCPO of model 2:",-sum(log(q1.m2.inla$cpo$cpo)),"\n")
```

```{r}
plot(airlines$year,q1.m2.inla$summary.fitted.values$sd,type="l",xlab="Year",ylab="sd(mu(t))")
```

```{r}
plot(airlines$year,q1.m2.inla$summary.fitted.values$mean,type="l",xlab="Year",ylab="E(mu(t))")
```

```{r}
#Contribution of random effect
plot(airlines$year,q1.m2.inla$summary.random$year1975$mean,type="l",xlab="Year",ylab="Posterior mean of rw1 temporal random effect")
```

**1.3. Fit a Poisson model on Fatal in INLA with Year as a covariate,
with offset set as log(miles) (see Workshop 4, question 1.5), using AR-1
random effect on the year, i.e. a Poisson model of the form**
$$\log(\mu(t))=\beta_0 + \beta_1 (t-\bar{t})+\log(\text{miles}(t))+f(t).$$**Compute
the DIC, NLSCPO, and WAIC scores. Plot the evolution of the posterior
standard deviation of the fitted values (**$\mu(t)$**) over the years
1976-2001.**

```{r}
prior.beta=list(mean.intercept = 0, prec.intercept = 0.1,
                    mean = 0, prec = 1)
q1.m3.inla <- inla(fatal ~ 1+scale(year1975,center = TRUE, scale = FALSE)+f(year1975,model="ar1"),family="Poisson",data=airlines,offset=log(miles), control.fixed=prior.beta,control.compute=list(cpo=T,dic=T,waic=T))
summary(q1.m3.inla)
```

```{r}
cat("Marginal log-likelihood of model 3:",q1.m3.inla$mlik[1],"\n")
cat("DIC of model 3:",q1.m3.inla$dic$dic,"\n")
cat("WAIC of model 3:",q1.m3.inla$waic$waic,"\n")
cat("NSLCPO of model 3:",-sum(log(q1.m3.inla$cpo$cpo)),"\n")
```

```{r}
plot(airlines$year,q1.m3.inla$summary.fitted.values$sd,type="l",xlab="Year",ylab="sd(mu(t))")
```

```{r}
plot(airlines$year,q1.m3.inla$summary.fitted.values$mean,type="l",xlab="Year",ylab="E(mu(t))")
```

```{r}
#Contribution of random effect
plot(airlines$year,q1.m3.inla$summary.random$year1975$mean,type="l",xlab="Year",ylab="Posterior mean of ar-1 temporal random effect")
```

**1.4. Fit a Poisson model on Fatal in INLA with Year as a covariate,
with offset set as log(miles) (see Workshop 4, question 1.5), using AR-2
random effect on the year, i.e. a Poisson model of the form**
$$\log(\mu(t))=\beta_0 + \beta_1 (t-\bar{t})+\log(\text{miles}(t))+f(t).$$**Compute
the DIC, NLSCPO, and WAIC scores. Plot the evolution of the posterior
standard deviation of the fitted values (**$\mu(t)$**) over the years
1976-2001. Discuss and compare all 4 models in 1.1. - 1.4.**

```{r}
prior.beta=list(mean.intercept = 0, prec.intercept = 0.1,
                    mean = 0, prec = 1)
q1.m4.inla <- inla(fatal ~ 1+scale(year1975,center = TRUE, scale = FALSE)+f(year1975,model="ar",order=2),family="Poisson",data=airlines,offset=log(miles), control.fixed=prior.beta,control.compute=list(cpo=T,dic=T,waic=T),control.predictor =list(compute=TRUE))
summary(q1.m4.inla)
```

```{r}
cat("Marginal log-likelihood of model 4:",q1.m4.inla$mlik[1],"\n")
cat("DIC of model 4:",q1.m4.inla$dic$dic,"\n")
cat("WAIC of model 4:",q1.m4.inla$waic$waic,"\n")
cat("NSLCPO of model 4:",-sum(log(q1.m4.inla$cpo$cpo)),"\n")
```

As we can see, model 4 is the best amongst all four models considered in
terms of DIC, WAIC and NLSCPO.

```{r}
plot(airlines$year,q1.m4.inla$summary.fitted.values$sd,type="l",xlab="Year",ylab="sd(mu(t))")
```

```{r}
plot(airlines$year,q1.m4.inla$summary.fitted.values$mean,type="l",xlab="Year",ylab="E(mu(t))")
```

```{r}
#Contribution of random effect
plot(airlines$year,q1.m4.inla$summary.random$year1975$mean,type="l",xlab="Year",ylab="Posterior mean of ar-2 temporal random effect")
```

As we can see, the posterior mean E(mu(t)) in AR-2 model varies more
smoothly in terms of Year than in the previous models, but sd(mu(t)) is
significantly higher (2.2-3) compared to the other models.

**1.5. Re-fit all 4 models from 1.1 - 1.4 using only data from
1976-1997, and use these to predict the number of incidents in 1998,
1999, 2000 and 2001 (Hint: include these years as NA in the response).
Compute the posterior predictive mean number of incidents each year from
1998-2001 and compute the differences between the predicted and actual
values. Discuss the predictive accuracy of the models.**

```{r}
airlines.pred=airlines
n=nrow(airlines)
airlines.pred$fatal[(n-3):n]=NA

q1.m1.inla.pred <- inla(fatal ~ 1+scale(year1975,center = TRUE, scale = FALSE),family="Poisson",data=airlines.pred,offset=log(miles), control.fixed=prior.beta, control.predictor = list(link=1))
q1.m1.inla.pred$summary.fitted.values$mean[(n-3):n]
airlines$fatal[(n-3):n]
diff=q1.m1.inla.pred$summary.fitted.values$mean[(n-3):n]-airlines$fatal[(n-3):n]
diff
mean(abs(diff))
```

```{r}
q1.m2.inla.pred <- inla(fatal ~ 1+scale(year1975,center = TRUE, scale = FALSE)+f(year1975,model="rw1"),family="Poisson",data=airlines.pred,offset=log(miles), control.fixed=prior.beta, control.predictor = list(link=1))
q1.m2.inla.pred$summary.fitted.values$mean[(n-3):n]
airlines$fatal[(n-3):n]
diff=q1.m2.inla.pred$summary.fitted.values$mean[(n-3):n]-airlines$fatal[(n-3):n]
diff
mean(abs(diff))
```

```{r}
q1.m3.inla.pred <- inla(fatal ~ 1+scale(year1975,center = TRUE, scale = FALSE)+f(year1975,model="ar1"),family="Poisson",data=airlines.pred,offset=log(miles), control.fixed=prior.beta, control.predictor = list(link=1))
q1.m3.inla.pred$summary.fitted.values$mean[(n-3):n]
airlines$fatal[(n-3):n]
diff=q1.m3.inla.pred$summary.fitted.values$mean[(n-3):n]-airlines$fatal[(n-3):n]
diff
mean(abs(diff))
```

```{r}
q1.m4.inla.pred <- inla(fatal ~ 1+scale(year1975,center = TRUE, scale = FALSE)+f(year1975,model="ar",order=2),family="Poisson",data=airlines.pred,offset=log(miles), control.fixed=prior.beta, control.predictor = list(link=1))
q1.m4.inla.pred$summary.fitted.values$mean[(n-3):n]
airlines$fatal[(n-3):n]
diff=q1.m4.inla.pred$summary.fitted.values$mean[(n-3):n]-airlines$fatal[(n-3):n]
diff
mean(abs(diff))
```

As we can see, the AR-2 model slightly lower prediction error on these 4
years. Combined with the better DIC, WAIC and NLSCPO scores, it
indicates that this is likely the best model in terms of predictive
performance. Nevertheless, the size of the test data (4 observations)
and the training data (22 observations) are too small to make a
definitive conclusion.

# 2. **Modelling rainfall data from Brazil**

**Our dataset contains daily rainfall amounts (in mm) recorded at 616
stations in the 2011 year in Paraná state, Brazil.**

```{r}
require(splancs) 
require(INLA) 
require(gridExtra) 
require(lattice) 

data(PRprec) 
head(PRprec)
```

```{r}
table(rowSums(is.na(PRprec[, 3 + 1:31])))
```

As we can see, there are some missing values in the dataset. We compute
the total rainfall in January and save it as a new column called
precJanuary (missing values are ignored using na.rm=TRUE).

```{r}
PRprec$precJanuary <- rowSums(PRprec[, 3 + 1:31], na.rm = TRUE)
meanAltitude=mean(PRprec$Altitude, na.rm=TRUE)
PRprec$Altitude[is.na(PRprec$Altitude)]=meanAltitude
```

**2.1. Using INLA, implement a Bayesian linear regression model with
Gaussian likelihood for the total rainfall in January (precJanuary) with
Altitude, Latitude and Longitude as covariates. Interpret the summary of
the fitted model. Compute the DIC, NLSCPO and WAIC scores.**

```{r}
prior.beta=list(mean.intercept = 0, prec.intercept = 1e-6,
                    mean = 0, prec = 1e-4)
q2.m1.inla <- inla(precJanuary ~ 1+scale(Altitude)+scale(Latitude)+scale(Longitude),family="Gaussian",data=PRprec, control.fixed=prior.beta,control.compute=list(cpo=T,dic=T,waic=T))
summary(q2.m1.inla)
```

**2.2. In this question, we will build a spatial model.**

```{r}
coords <- as.matrix(PRprec[, 1:2]) 
plot(coords)
```

```{r}
prdomain <- inla.nonconvex.hull(as.matrix(PRprec[, 1:2]),
  convex = -0.03, concave = -0.05,
  resolution = c(100, 100))
plot(prdomain)
```

```{r}
prmesh <- inla.mesh.2d(boundary = prdomain,
  max.edge = c(0.45, 1), cutoff = 0.2)
plot(prmesh)
```

**Using INLA or inlabru, implement a Bayesian regression model (Gaussian
likelihood) for the total rainfall in January (precJanuary) with
Altitude, Latitude and Longitude as fixed effects, an SPDE random effect
for the location (Longitude/Latitude). Interpret the summary of the
fitted model. Plot the posterior mean of the spatial random effect.
Compute the DIC, NLSCPO and WAIC scores.**

```{r}
Locations = cbind(PRprec$Longitude, PRprec$Latitude) 
loc.spde = inla.spde2.pcmatern(mesh = prmesh,
           prior.range = c(1, 0.1),
           prior.sigma = c(200, 0.1))
#The choices of the priors are crucial for good performance
#See https://rdrr.io/github/INBO-BMK/INLA/man/inla.spde2.pcmatern.html for a description on the meaning of the prior parameters
#In particular, we need to set prior.sigma to sufficiently large, since the response variable changes on a scale of 200.

loc.A <- inla.spde.make.A(prmesh, loc = Locations) 
loc.w <- inla.spde.make.index('w', n.spde = loc.spde$n.spde) 

# Make the model matrix using the final model selection formula without response and intercept.
X0 <- model.matrix(as.formula(" ~ 0+Altitude+Longitude+Latitude"), data = PRprec) 
X <- as.data.frame(X0) # convert to a data frame. 

# Making the stack ####

N <- nrow(PRprec)


StackPR<- inla.stack(
  data = list(y = PRprec$precJanuary), # specify the response variable
  
  A = list(1, 1, loc.A), # Vector of Multiplication factors for  fixed effects              
  
  effects = list(
    Intercept = rep(1, N), # specify the manual intercept!
    X = X, # attach the model matrix
   # insert vectors of any random effects
    w = loc.w) ) # attach the w 
```

```{r}
q2.m2.inla <- inla(y ~ 0 + Intercept +Altitude+Longitude+Latitude
            +f(w, model = loc.spde),     
            family = "Gaussian",
            data = inla.stack.data(StackPR),
            control.compute = list(cpo=T,dic = T,waic=T),
            control.predictor = list(A = inla.stack.A(StackPR)),control.inla=list(tolerance=1e-10))
summary(q2.m2.inla)

```

```{r}
library(INLA)
library(inlabru)

Locations = data.frame(easting=PRprec$Longitude, northing=PRprec$Latitude) 
loc.spde = inla.spde2.pcmatern(mesh = prmesh,
           prior.range = c(1, 0.5), 
           prior.sigma = c(200, 0.1)) 



PRprec$geometry <- sf::st_as_sf(Locations, coords = c("easting", "northing"))$geometry

cmp <- precJanuary ~ floc(geometry, model = loc.spde)+Longitude+Latitude+Altitude+ Intercept(1)


q2.m2.bru <- bru(components=cmp, family = "gaussian",data=PRprec,
                    samplers = prdomain,
    domain = list(coordinates = prmesh),
                 options=list(control.inla=list(tolerance=1e-10)))
summary(q2.m2.bru)
```

```{r}

if(!require(ggregplot)){
    devtools::install_github("gfalbery/ggregplot")
    library(ggregplot)
}
library(ggplot2) 
library(tidyverse)
library(RColorBrewer)


ggField(q2.m2.inla, prmesh, Groups = 1,Res=600) + scale_fill_brewer(palette = "RdYlBu")
  #scale_fill_gradientn(colours=brewer.pal(11, "RdYlBu"))
  
  #brewer.pal(11, "RdYlBu")#scale_fill_brewer(palette = "RdYlBu")


```

```{r}
library(INLA)
library(inlabru)
library(fmesher)
library(ggplot2)
library(sp)
library(RColorBrewer)


  pix <- fm_pixels(prmesh, dims = c(200, 200))

  pred <- predict(q2.m2.bru, pix, ~ floc)

  ggplot() + gg(pred, geom = "tile") +
  ggtitle("Posterior mean of spatial random effect")+
  scale_fill_gradientn(colours=brewer.pal(11, "RdYlBu"),limits=range(pred$mean))
```

**2.3. We define a new covariate related to the distance to the ocean.**

```{r}
coords <- as.matrix(PRprec[, 1:2]) 
mat.dists <- spDists(coords, PRborder[1034:1078, ],
  longlat = TRUE) 
PRprec$oceanDist <- apply(mat.dists, 1, min) 
```

**Using INLA or inlabru, implement a Bayesian regression model (Gaussian
likelihood) for the total rainfall in January (precJanuary) with
Altitude, Latitude and Longitude as fixed effects, an SPDE random effect
for the location (Longitude/Latitude), and additional random effects on
the Altitude, and distance to the ocean, oceanDist (these can be rw1,
ar1, or SPDE). Interpret the summary of the fitted model. Plot the
posterior mean of the spatial random effect. Compute the DIC, NLSCPO and
WAIC scores. Compare all 3 models in questions 2.1 - 2.3.**

```{r}
library(INLA)
library(inlabru)

Locations = data.frame(easting=PRprec$Longitude, northing=PRprec$Latitude) 
loc.spde = inla.spde2.pcmatern(mesh = prmesh, 
           prior.range = c(1, 0.5), 
           prior.sigma = c(200, 0.1)) 
a.mesh <- inla.mesh.1d(seq(min(PRprec$Altitude),max(PRprec$Altitude),length.out=100))
a.spde = inla.spde2.pcmatern(mesh = a.mesh, 
                             prior.range = c(100, 0.5), 
                             prior.sigma = c(1000, 0.1)) # Making SPDE

d.mesh <- inla.mesh.1d(seq(min(PRprec$oceanDist),max(PRprec$oceanDist),length.out=100))
d.spde = inla.spde2.pcmatern(mesh = d.mesh, 
                             prior.range = c(100, 0.5), 
                             prior.sigma = c(1000, 0.1)) # Making SPDE

PRprec$geometry <- sf::st_as_sf(Locations, coords = c("easting", "northing"))$geometry



cmp <- precJanuary ~ floc(geometry, model = loc.spde)+Longitude+Latitude+Altitude+ oceanDist+fa(Altitude,model=a.spde)+fo(oceanDist,model=d.spde)+Intercept(1)


q2.m3.bru <- bru(components=cmp, family = "gaussian",data=PRprec, options=list(control.inla=list(tolerance=1e-10)))
summary(q2.m3.bru)
```

```{r}
plot(d.mesh$loc,q2.m3.bru$summary.fixed$mean[4]*d.mesh$loc+q2.m3.bru$summary.random$fo$mean,xlab="distance from ocean",ylab="mean of total effect (fixed + random effect)",type="l")
```

```{r}
plot(a.mesh$loc,q2.m3.bru$summary.fixed$mean[3]*a.mesh$loc+q2.m3.bru$summary.random$fa$mean,xlab="altitude",ylab="mean of total effect (fixed + random effect)",type="l")
```

```{r}

  pix <- fm_pixels(prmesh, dims = c(200, 200))

  pred <- predict(q2.m3.bru, pix, ~ floc)

  ggplot() + gg(pred, geom = "tile") +
  ggtitle("Posterior mean of spatial random effect")+
  scale_fill_gradientn(colours=brewer.pal(11, "RdYlBu"),limits=range(pred$mean))
```

**2.4. Compute the total rainfall in all 12 months in the dataset, at
each location. Define a new data frame where the response column
monthlyprec contains the monthly rainfall, the month covariate denotes
the number of the month, and the Longitude, Latitude and Altitude
covariates are still included for each observation point. Using INLA or
inlabru, implement a Bayesian regression model (Gaussian likelihood) for
the monthly rainfall (monthlyprec) with Altitude, Latitude and Longitude
as fixed effects, an SPDE random effect for the location
(Longitude/Latitude), and additional random effects on the month,
Altitude, oceanDist covariates. Interpret the summary of the fitted
model. Plot the posterior mean of the spatial random effect. Compute the
DIC, NLSCPO and WAIC scores. Discuss and interpret the results.**

```{r}
n=nrow(PRprec)
PRprec.monthly<-data.frame(Latitude=PRprec$Latitude,Longitude=PRprec$Longitude,Altitude=PRprec$Altitude,
                           oceanDist=PRprec$oceanDist,geometry=PRprec$geometry,rain=0,month=0)
PRprec.monthly<-do.call("rbind", replicate(12, PRprec.monthly, simplify = FALSE))
strstart<-list("d01","d02","d03","d04","d05","d06","d07","d08","d09","d10","d11","d12")
for(m in 1:12){
PRprec.monthly$rain[((m-1)*n+1):(m*n)]<-rowSums(PRprec[,grepl(strstart[[m]],colnames(PRprec),fixed=TRUE)],na.rm=TRUE)
PRprec.monthly$month[((m-1)*n+1):(m*n)]=m
}
```

```{r}
library(INLA)
library(inlabru)

Locations = data.frame(easting=PRprec$Longitude, northing=PRprec$Latitude) 
loc.spde = inla.spde2.pcmatern(mesh = prmesh, 
           prior.range = c(1, 0.5), 
           prior.sigma = c(200, 0.1)) 
a.mesh <- inla.mesh.1d(seq(min(PRprec$Altitude),max(PRprec$Altitude),length.out=100))
a.spde = inla.spde2.pcmatern(mesh = a.mesh, 
                             prior.range = c(100, 0.5), 
                             prior.sigma = c(200, 0.1)) # Making SPDE

d.mesh <- inla.mesh.1d(seq(min(PRprec$oceanDist),max(PRprec$oceanDist),length.out=100))
d.spde = inla.spde2.pcmatern(mesh = d.mesh, 
                             prior.range = c(100, 0.5), 
                             prior.sigma = c(200, 0.1)) # Making SPDE

PRprec$geometry <- sf::st_as_sf(Locations, coords = c("easting", "northing"))$geometry



cmp <- rain ~ floc(geometry, model = loc.spde)+Longitude+Latitude+Altitude+ oceanDist+fa(Altitude,model=a.spde)+fo(oceanDist,model=d.spde)+f(month,model="rw1")+Intercept(1)

q2.m4.bru <- bru(components=cmp, family = "gaussian",data=PRprec.monthly, options=list(control.inla=list(tolerance=1e-10)))
summary(q2.m4.bru)
```

```{r}

  pix <- fm_pixels(prmesh, dims = c(200, 200))

  pred <- predict(q2.m4.bru, pix, ~ floc)

  ggplot() + gg(pred, geom = "tile") +
  ggtitle("Posterior mean of spatial random effect")+
  scale_fill_gradientn(colours=brewer.pal(11, "RdYlBu"),limits=range(pred$mean))
```

```{r}
plot(d.mesh$loc,q2.m4.bru$summary.fixed$mean[4]*d.mesh$loc+q2.m4.bru$summary.random$fo$mean,xlab="distance from ocean",ylab="mean of total effect (fixed + random effect)",type="l")
```

The areas immediately near the ocean seem to be quite rainy. Further
away, the situation is less clear - this could be due to some
collinearity between altitude (which in general increases rainfall) and
distance from the ocean.

```{r}
plot(a.mesh$loc,q2.m4.bru$summary.fixed$mean[3]*a.mesh$loc+q2.m4.bru$summary.random$fa$mean,xlab="altitude",ylab="mean of total effect (fixed + random effect)",type="l")
```

As expected, higher altitude seems to result in more rain.

```{r}
plot(1:12,q2.m4.bru$summary.random$f$mean,xlab="month",ylab="mean of random effect",type="l")
```

As Encyclopedia Britannica states, "The basins of the Alto Paraná and
Paraguay have a hot and humid climate throughout the year. The winters
(April to September) are dry, and the summers (October to March) are
rainy." This is more-or-less consistent with what we have here, the
difference in some month could be due to year-to-year variations (we
only used 1 year of data).
