---
editor_options:
  markdown:
    wrap: 72
output: pdf_document
---

**University of Edinburgh**\
**School of Mathematics**\
**Bayesian Data Analysis, 2023/2024, Semester 2**\

**Workshop 5: Temporal and Spatial Random Effects**

```{r}
library(INLA)
```

# 1. **Modelling fatal airline accidents from 1976 through 2001.**

**This exercise has been taken largely from a shortcourse at the
University of Copenhagen which occurred in January 2013 and notes from
Gurrin, Carstensen, Hojsgaard, and Ekstrom. The dataset `airline.RData`
is available on Learn.**

**The fields are:**\
**- Year1975 (number of years after 1975),**

**- Year,**

**- Fatal (number of fatal airline accidents),**

**- Miles (total passenger miles, in** $10^{11}$ **miles, e.g.,**
$3.863 = 3.683*10^{11} \text{miles} = 368.3$ **Billion miles),**

**- Rate (fatalities per** $10^{11}$ **passenger miles).**

**You will be fitting some temporal random effect Poisson models to
Fatal.**

```{r}
load("airline.RData")
head(airlines)
```

**1.1. Fit a simple Poisson model on Fatal in INLA with Year as a
covariate, with offset set as log(miles) (see Workshop 4, question 1.5),
i.e. a Poisson model of the form**
$$\log(\mu(t))=\beta_0 + \beta_1 (t-\bar{t})+\log(\text{miles}(t))$$
**Here miles is the total number of miles of flights per year divided by
10\^11 (ranging from 5 to 20 during the period 1976-2001.**

**Either choose your own priors for** $\beta_0$ and $\beta_1$, or use
the default priors.

**Compute the DIC, NLSCPO, and WAIC scores (WAIC can be computed by
setting waic=T in control.compute).**

```{r}
airlines$year.ctr = airlines$year1975 - mean(airlines$year1975)
prior.beta = list(mean.intercept=0, prec.intercept=0.5,
                  mean=0, prec=0.5)

air.ord = inla(fatal ~ year.ctr, data=airlines, family="poisson",
               offset=log(miles), control.fixed=prior.beta,
               control.compute=list(config=T, cpo=T, dic=T, waic=T))

air.ord$dic$dic
-sum(log(air.ord$cpo$cpo))
air.ord$waic$waic
```

**1.2. Fit a Poisson model on Fatal in INLA with Year as a covariate,
with offset set as log(miles) (see Workshop 4, question 1.5), using RW1
random effect on the year, i.e. a Poisson model of the form**
$$\log(\mu(t))=\beta_0 + \beta_1 (t-\bar{t})+\log(\text{miles}(t))+f(t).$$**Compute
the DIC, NLSCPO, and WAIC scores. Plot the evolution of the posterior
standard deviation of the fitted values (**$\mu(t)$**) over the years
1976-2001.**

```{r}
air.rw = inla(fatal ~ 1 + year.ctr + f(year1975, model="rw1"), data=airlines,
              family="poisson", offset=log(miles),
              control.fixed=prior.beta,
              control.compute=list(config=T, cpo=T, dic=T, waic=T))

# when there is no NA values, summary.fitted.values are mu instead of eta by default, no need of predictor & link=1

# cannot use year.ctr as the random walk indicator!
# constr = TRUE?

air.rw$dic$dic
-sum(log(air.rw$cpo$cpo))
air.rw$waic$waic

plot(airlines$year, air.rw$summary.fitted.values$mean)
plot(airlines$year, air.rw$summary.fitted.values$sd)
```

**1.3. Fit a Poisson model on Fatal in INLA with Year as a covariate,
with offset set as log(miles) (see Workshop 4, question 1.5), using AR-1
random effect on the year, i.e. a Poisson model of the form**
$$\log(\mu(t))=\beta_0 + \beta_1 (t-\bar{t})+\log(\text{miles}(t))+f(t).$$**Compute
the DIC, NLSCPO, and WAIC scores. Plot the evolution of the posterior
standard deviation of the fitted values (**$\mu(t)$**) over the years
1976-2001.**

```{r}
air.ar = inla(fatal ~ 1 + year.ctr + f(year1975, model="ar1"), data=airlines,
              family="poisson", offset=log(miles),
              control.fixed=prior.beta,
              control.compute=list(config=T, cpo=T, dic=T, waic=T))

air.ar$dic$dic
-sum(log(air.ar$cpo$cpo))
air.ar$waic$waic

plot(airlines$year, air.ar$summary.fitted.values$mean)
plot(airlines$year, air.ar$summary.fitted.values$sd)
```

**1.4. Fit a Poisson model on Fatal in INLA with Year as a covariate,
with offset set as log(miles) (see Workshop 4, question 1.5), using AR-2
random effect on the year, i.e. a Poisson model of the form**
$$\log(\mu(t))=\beta_0 + \beta_1 (t-\bar{t})+\log(\text{miles}(t))+f(t).$$**Compute
the DIC, NLSCPO, and WAIC scores. Plot the evolution of the posterior
standard deviation of the fitted values (**$\mu(t)$**) over the years
1976-2001. Discuss and compare all 4 models in 1.1. - 1.4.**

```{r}
air.ar2 = inla(fatal ~ 1 + year.ctr + f(year1975, model="ar", order=2),
               data=airlines, family="poisson", offset=log(miles),
               control.fixed=prior.beta,
               control.compute=list(config=T, cpo=T, dic=T, waic=T))

air.ar2$dic$dic
-sum(log(air.ar2$cpo$cpo))
air.ar2$waic$waic

plot(airlines$year, air.ar2$summary.fitted.values$mean)
plot(airlines$year, air.ar2$summary.fitted.values$sd)
```

**1.5. Re-fit all 4 models from 1.1 - 1.4 using only data from
1976-1997, and use these to predict the number of incidents in 1998,
1999, 2000 and 2001 (Hint: include these years as NA in the response).
Compute the posterior predictive mean number of incidents each year from
1998-2001 and compute the differences between the predicted and actual
values. Discuss the predictive accuracy of the models.**

```{r}
airlines.pred = airlines
airlines.pred$fatal[23:26] = NA
airlines.pred$year.ctr = airlines$year1975 - mean(airlines$year1975[1:22])

air.ord = inla(fatal ~ year.ctr, data=airlines.pred, family="poisson",
               offset=log(miles), control.fixed=prior.beta,
               control.predictor=list(compute=T, link=1),
               control.compute=list(config=T, cpo=T, dic=T, waic=T))

# "compute the posterior mean number of ..., for mean we do not need to
#   sample and calculate mean through rpois, since the link=1 output, lambda,
#   is exactly the expected number of accidents.
fatal.true = airlines$fatal[23:26]

fatal.pred.mean.ord = air.ord$summary.fitted.values[23:26, "mean"]
mean(abs(fatal.pred.mean.ord - fatal.true))

air.rw = inla(fatal ~ 1 + year.ctr + f(year1975, model="rw1"), data=airlines.pred,
              family="poisson", offset=log(miles),
              control.fixed=prior.beta,
               control.predictor=list(compute=T, link=1),
              control.compute=list(config=T, cpo=T, dic=T, waic=T))

fatal.pred.mean.rw = air.rw$summary.fitted.values[23:26, "mean"]
mean(abs(fatal.pred.mean.rw - fatal.true))

air.ar = inla(fatal ~ 1 + year.ctr + f(year1975, model="ar1"), data=airlines.pred,
              family="poisson", offset=log(miles),
              control.fixed=prior.beta,
               control.predictor=list(compute=T, link=1),
              control.compute=list(config=T, cpo=T, dic=T, waic=T))

fatal.pred.mean.ar = air.ar$summary.fitted.values[23:26, "mean"]
mean(abs(fatal.pred.mean.ar - fatal.true))

air.ar2 = inla(fatal ~ 1 + year.ctr + f(year1975, model="ar", order=2),
               data=airlines.pred,
               family="poisson", offset=log(miles),
               control.fixed=prior.beta,
               control.predictor=list(compute=T, link=1),
               control.compute=list(config=T, cpo=T, dic=T, waic=T))

fatal.pred.mean.ar2 = air.ar2$summary.fitted.values[23:26, "mean"]
mean(abs(fatal.pred.mean.ar2 - fatal.true))
```

# 2. **Modelling rainfall data from Brazil**

**Our dataset contains daily rainfall amounts (in mm) recorded at 616
stations in the 2011 year in Paraná state, Brazil.**

```{r}
require(splancs) 
require(INLA) 
require(gridExtra) 
require(lattice) 

data(PRprec) 
head(PRprec)
```

```{r}
table(rowSums(is.na(PRprec[, 3 + 1:31])))
```

As we can see, there are some missing values in the dataset. We compute
the total rainfall in January and save it as a new column called
precJanuary (missing values are ignored using na.rm=TRUE).

```{r}
PRprec$precJanuary <- rowSums(PRprec[, 3 + 1:31], na.rm = TRUE)
meanAltitude=mean(PRprec$Altitude, na.rm=TRUE)
PRprec$Altitude[is.na(PRprec$Altitude)]=meanAltitude
```

**2.1. Using INLA, implement a Bayesian linear regression model with
Gaussian likelihood for the total rainfall in January (precJanuary) with
Altitude, Latitude and Longitude as covariates. Interpret the summary of
the fitted model. Compute the DIC, NLSCPO and WAIC scores.**

```{r}
prior.beta = list(mean.intercept=0, prec.intercept=1e-4,
                  mean=0, prec=1e-6)
rain.ord = inla(precJanuary ~ scale(Altitude) + scale(Latitude) + scale(Longitude),
                data=PRprec, family="gaussian", control.fixed=prior.beta,
                control.compute=list(dic=T, cpo=T, waic=T))

summary(rain.ord)

rain.ord$dic$dic
-sum(log(rain.ord$cpo$cpo))
rain.ord$waic$waic
```

**2.2. In this question, we will build a spatial model.**

```{r}
coords <- as.matrix(PRprec[, 1:2]) 
plot(coords)
```

```{r}
prdomain <- inla.nonconvex.hull(as.matrix(PRprec[, 1:2]),
  convex = -0.03, concave = -0.05,
  resolution = c(100, 100))
plot(prdomain)
```

```{r}
prmesh <- inla.mesh.2d(boundary = prdomain,
  max.edge = c(0.45, 1), cutoff = 0.2)
plot(prmesh)
```

**Using INLA or inlabru, implement a Bayesian regression model (Gaussian
likelihood) for the total rainfall in January (precJanuary) with
Altitude, Latitude and Longitude as fixed effects, an SPDE random effect
for the location (Longitude/Latitude). Interpret the summary of the
fitted model. Plot the posterior mean of the spatial random effect.
Compute the DIC, NLSCPO and WAIC scores.**

```{r}

Locations = cbind(PRprec$Longitude, PRprec$Latitude)

loc.a = inla.spde.make.A(prmesh, loc=Locations)
loc.spde = inla.spde2.pcmatern(mesh=prmesh,
                               prior.range=c(1, 0.1),
                               prior.sigma=c(200, 0.1))  # how to set this prior?
loc.w = inla.spde.make.index('w', n.spde=loc.spde$n.spde)

X0 = model.matrix( ~ 0 + Altitude + Latitude + Longitude, data=PRprec)
X = as.data.frame(X0)
N = nrow(PRprec)

StackPRprec = inla.stack(
  data=list(y=PRprec$precJanuary),
  A=list(1, 1, loc.a),
  effects=list(
    Intercept=rep(1, N),
    X=X,
    w=loc.w
  )
)


rain.spde = inla(y ~ 0 + Intercept + Altitude + Latitude + Longitude + f(w, model=loc.spde),
                 family="gaussian", data=inla.stack.data(StackPRprec),
                 control.compute=list(cpo=T, dic=T),
                 control.predictor=list(A=inla.stack.A(StackPRprec)),
                 control.inla=list(tolerance=1e-10))

summary(rain.spde)
```

```{r}
if(!require(ggregplot)){
    devtools::install_github("gfalbery/ggregplot")
    library(ggregplot)
}
library(ggplot2) 
library(tidyverse)
library(RColorBrewer)


ggField(rain.spde, prmesh, Groups = 1, Res=600) + scale_fill_brewer(palette = "RdYlBu")

rain.spde$dic$dic
-sum(log(rain.spde$cpo$cpo))
rain.spde$waic$waic
```

**2.3. We define a new covariate related to the distance to the ocean.**

```{r}
coords <- as.matrix(PRprec[, 1:2]) 
mat.dists <- spDists(coords, PRborder[1034:1078, ],
  longlat = TRUE) 
PRprec$oceanDist <- apply(mat.dists, 1, min) 
```

**Using INLA or inlabru, implement a Bayesian regression model (Gaussian
likelihood) for the total rainfall in January (precJanuary) with
Altitude, Latitude and Longitude as fixed effects, an SPDE random effect
for the location (Longitude/Latitude), and additional random effects on
the Altitude, and distance to the ocean, oceanDist (these can be rw1,
ar1, or SPDE). Interpret the summary of the fitted model. Plot the
posterior mean of the spatial random effect. Compute the DIC, NLSCPO and
WAIC scores. Compare all 3 models in questions 2.1 - 2.3.**

```{r}

```

**2.4. Compute the total rainfall in all 12 months in the dataset, at
each location. Define a new data frame where the response column
monthlyprec contains the monthly rainfall, the month covariate denotes
the number of the month, and the Longitude, Latitude and Altitude
covariates are still included for each observation point. Using INLA or
inlabru, implement a Bayesian regression model (Gaussian likelihood) for
the monthly rainfall (monthlyprec) with Altitude, Latitude and Longitude
as fixed effects, an SPDE random effect for the location
(Longitude/Latitude), and additional random effects on the month,
Altitude, oceanDist covariates. Interpret the summary of the fitted
model. Plot the posterior mean of the spatial random effect. Compute the
DIC, NLSCPO and WAIC scores. Discuss and interpret the results.**

```{r}

```
