---
editor_options: 
  markdown: 
    wrap: 72
---

**University of Edinburgh**

**School of Mathematics**

**Bayesian Data Analysis, 2023/2024, Semester 2**

**Workshop 2: Introduction to JAGS**

**Note**: Before starting this practical, you might want to spend some
time looking at the JAGS examples we have discussed during Lecture 3. If
you already read them, then go directly to question 1. The code below
loads JAGS.

```{r}
library(rjags)
#If it ran correctly, you should see 
#Loading required package: coda
#Linked to JAGS 4.3.1
#Loaded modules: basemod,bugs
```

**1. Simple linear regression with robustification**

**Winning Olympic Men's Long Jump Distances (adapted from Witmer,
2017)**

**The data are the winning men's long jump distances (m) from 1900
through 2008. You will fit a linear regression of the distances as a
function of Olympic year:** $$\begin{aligned}
Jump & = & \beta_0 + \beta_1 \mbox{Year} + \epsilon\end{aligned}$$
**three different ways: standard frequentist approach, a Bayesian
approach assuming normal errors, and a Bayesian approach assuming a**
$t$ **distribution for errors.**

**Run the following commands in `R` to begin (this will install and load
the package Stat2Data and load the Long Jump dataset).**

```{r}
require("Stat2Data")
#If you don't have this, run install.packages("Stat2Data")
library(Stat2Data)
data("LongJumpOlympics")   #Makes the dataset available in this R session
Jump <- LongJumpOlympics$Gold
Year <- LongJumpOlympics$Year
```

**1.1. We start by carrying out some exploratory data analysis. Plot
Jump vs Year. What does the relationship look like?**

```{r}
plot(Year, Jump, main="Winning men's long jump distance for each year")
```

**1.2. Fit a simple linear regression on Jump against Year using the
`lm` function, and make a plot of the data with the fitted line overlaid
using the `abline` function.**

```{r}
model.l <- lm(Jump ~ Year)
names(model.l)
summary(model.l)
plot(Year, Jump, main="Winning men's long jump distance for each year", pch=19, col="grey40")
abline(coef=model.l$coefficients, col="red")
```

**1.3. Based on this model, every 4 years we would expect the jump
distance to change by what amount?**

```{r}
cat("The changing amount of distance every 4 years will be: ", 4 * model.l$coefficients[2])
```

**1.4. Plot the residuals against Year (using the `residuals` function).
One year stands out, which one is it?**

```{r}
resid <- resid(model.l)
resid
plot(Year, resid, main="residuals vs. Year", pch=19, col="grey40")
abline(h=0)
```

**1.5. For a more detailed residual analysis, type `par(mfrow=c(2,2))`,
and use the `plot` function operating on the `lm` object (you'll see 4
plots).**

```{r}
par(mfrow=c(2, 2))
plot(model.l)
```

**1.6. Remove the outlier from the data set and refit the model, then
recompute the above residual diagnostics. What do you observe?**

```{r}
outlier.ind <- which(resid == max(resid))
Jump.ro <- Jump[-outlier.ind]
Year.ro <- Year[-outlier.ind]
model.l.ro <- lm(Jump.ro ~ Year.ro)
summary(model.l.ro)

resid.ro <- residuals(model.l.ro)
plot(Year.ro, resid.ro, main="residuals vs. Year", pch=19, col="grey40")
abline(h=0)

par(mfrow=c(2, 2))
plot(model.l.ro)
```

**1.7. Carry out a Bayesian linear regression analysis using `rjags`. As
in the frequentist case assume** $\epsilon$ $\sim$ Normal(0,
$\sigma^2$). Use the following priors for the three parameters:
$$\begin{aligned}
\beta_0, \beta_1 & \sim & \mbox{Normal} \left ( \mu_0=0, \tau_0=0.001 \right ) \\
\tau & \sim & \mbox{Gamma} \left ( a=0.1, b=0.1 \right )
\end{aligned}$$

**Write the *model* statement, which includes the likelihood calculation
and the prior distribution. Include a calculation of**
$\sigma = 1/\sqrt{\tau}$.

```{r}
model_string <- "
model {
  b0 ~ dnorm(mu0, tau0)
  b1 ~ dnorm(mu0, tau0)
  
  tau ~ dgamma(a, b)
  
  sigma = 1 / sqrt(tau)
  
  for (i in 1: n) {
    mu[i] = b0 + b1 * Year[i]
    Jump[i] ~ dnorm(mu[i], tau)
  }
}
"
```

**1.8. Create an `R` object for the data, which includes Jump, Year,**
$n$**=26 and the values of the prior hyperparameters** $\mu_0$**,**
$\tau_0$**,** $a$ **and** $b$**.**

```{r}
data <- list(Jump=Jump, Year=Year, n=26, mu0=0, tau0=0.001, a=0.1, b=0.1)
```

**1.9. Create an `R` object for 3 sets of initial values; e.g.,**

```{r}
my.inits <- list(list(b0=0.1,b1=0.2,tau=0.1),
           list(b0=-1,b1=3,tau=0.3),
           list(b0=1,b1=0,tau=.8))
```

**1.10. Execute `jags.model` using the above objects. Note `n.chains`
should be set equal to 3. How many unobserved stochastic nodes were
there? How many observed?**

```{r}
model.j <- jags.model(textConnection(model_string), data=data, n.chains=3, inits=my.inits)
```

**1.11. Use `update` to carry out an initial MCMC run (burn-in) with
1,000 iterations.**

```{r}
update(model.j, n.iter=1000)
```

**1.12. Now make a longer MCMC run using the `coda.samples` function
with 10,000 iterations and have the results for** $\beta_0$, $\beta_1$,
and $\sigma$ returned.

```{r}
res <- coda.samples(model.j, variable.names=c("b0", "b1", "sigma"), n.iter=10000)
```

**1.13. Plot the results from `coda.samples`. These are the trace plots.
Do you think that the chains have converged for each of the 3
parameters?**

```{r}
plot(res)
```

**1.14. You may have noticed from the trace plots that** $\beta_0$
**and** $\beta_1$ **are mixing slowly. That's indicative of significant
autocorrelation. Use the `acf` function to see how much correlation
there is. For example, if the results from `coda.samples` are called
`res`, for a parameter named beta0:, you can write
`acf(res[[1]][,"beta0"],lag.max=100)`.**

```{r}
par(mfrow=c(1, 2))
acf(res[[1]][, "b0"], lag.max=100)
acf(res[[1]][, "b1"], lag.max=100)
# acf(res[[1]][, "sigma"], lag.max=100)
```

**1.15. Also take a look at the effective sample sizes per parameter,
e.g.,`effectiveSize(res[[1]][,"beta0"])`**

```{r}
effectiveSize(res[[1]][, "b0"])
effectiveSize(res[[1]][, "b1"])
effectiveSize(res[[1]][, "sigma"])
```

**1.16. In Lecture 3, the Gelman-Rubin (Brooks-Gelman-Rubin) statistic
was discussed. This is a quantitative measure of apparent convergence
that is based upon the degree of overlap of 2 or more chains after each
iteration. The BGR statistic roughly approximates the ratio of the
variability between chains to the variability within chains (like an F
statistic in ANOVA). The general idea of the statistic is that the the
ratio of those two measures should be around 1 at convergence, thus
BGR=1 is "good". Use the `coda` package function called `gelman.plot` to
plot the BGR statistic for each of the parameters against the MCMC
iteration. And use `gelman.diag` for numerical summaries. What do you
think about convergence now?**

```{r}
gelman.plot(res)
gelman.diag(res)
```

**1.17. Centring the covariate, in this case Year, sometimes helps
convergence. Modify your Model statement slightly by creating a new
variable `meanY`, and then subtract that from the `Year[i]` values in
the for loop. Repeat the above steps. How does convergence now look? Use
the `summary` function on the JAGS output to examine the posterior means
and standard deviations for** $\beta_0$**,** $\beta_1$**, and**
$\sigma$**. How do the posterior mean for** $\beta_1$ **and** $\sigma$
**compare to the maximum likelihood estimates obtained in 1.2?**

```{r}
model_string2 <- "
model {
  b0 ~ dnorm(mu0, tau0)
  b1 ~ dnorm(mu0, tau0)
  
  tau ~ dgamma(a, b)
  
  sigma = 1 / sqrt(tau)
  
  meanY = mean(Year)
  
  for (i in 1: n) {
    mu[i] = b0 + b1 * (Year[i] - meanY)
    Jump[i] ~ dnorm(mu[i], tau)
  }
}
"

data <- list(Jump=Jump, Year=Year, n=26, mu0=0, tau0=0.001, a=0.1, b=0.1)
model.j.cent <- jags.model(textConnection(model_string2), data=data, inits=my.inits, n.chains=3)
update(model.j.cent, n.iter=1000)
res.cent <- coda.samples(model.j.cent, variable.names=c("b0", "b1", "sigma"), n.iter=10000)
```

```{r}
plot(res.cent)
par(mfrow=c(1, 2))
acf(res.cent[[1]][, "b0"], lag.max=100)
acf(res.cent[[1]][, "b1"], lag.max=100)
effectiveSize(res.cent[[1]][, "b0"])
effectiveSize(res.cent[[1]][, "b1"])
gelman.plot(res.cent)
gelman.diag(res.cent)
# better results, but still need more burn-in
```

**1.18. *Robustifying the regression.* As was noted in Lecture 3, the
effects of extreme observations or "outliers" on regressing results can
be diminished by using a** $t$ **distribution for the observations. For
simplicity, assume a** $t$ **distribution with 3 df for the distribution
of errors. Revise the JAGS model code accordingly (continuing to work
with the centred covariate) and re-run. Recall from Lecture 3 that the
necessary change to the code is to replace `dnorm` with `dt` and add an
additional argument to `data` for the df (=3). How did the posterior
mean of** $\beta_1$ **change? Compare it to the estimate in 1.2 when the
extreme observation is removed.**

-   

    ```{r}
    model_string3 <- "
    model {
      b0 ~ dnorm(mu0, tau0)
      b1 ~ dnorm(mu0, tau0)
      
      tau ~ dgamma(a, b)
      
      sigma = 1 / sqrt(tau)
      
      meanY = mean(Year)
      
      for (i in 1: n) {
        mu[i] = b0 + b1 * (Year[i] - meanY)
        Jump[i] ~ dt(mu[i], tau, nu)
      }
    }
    "

    data <- list(Jump=Jump, Year=Year, n=26, mu0=0, tau0=0.001, a=0.1, b=0.1, nu=3)
    model.j.rob <- jags.model(textConnection(model_string3), data=data, inits=my.inits, n.chains=3)
    update(model.j.rob, n.iter=1000)
    res.rob <- coda.samples(model.j.rob, variable.names=c("b0", "b1", "sigma"), n.iter=10000)
    ```

    ```{r}
    summary(res.rob)
    plot(res.rob)
    par(mfrow=c(1, 2))
    acf(res.rob[[1]][, "b0"], lag.max=100)
    acf(res.rob[[1]][, "b1"], lag.max=100)
    effectiveSize(res.rob[[1]][, "b0"])
    effectiveSize(res.rob[[1]][, "b1"])
    gelman.plot(res.rob)
    gelman.diag(res.rob)
    ```

**2. Nonlinear Regression. Newton's law of cooling, from Bates and Watts
(2008)**

**The following data are measurements over a 41 minute period of the
temperature of a bore after that bore had been rubbed inside "a
stationary cylinder and pressed against the bottom by means of a screw".
The bore was turned by a team of horses (this is an experiment with
friction from 1798 by a Count Rumford).**

```{r}
#minutes
elapsed.time <- c(4,5,7,12,14,16,20,24,28,31,34,37.5,41)
#Fahrenheit
temperature <- c(126,125,123,120,119,118,116,115,114,113,112,111,110) 
```

**An underlying theoretical model based on Newton's law of cooling
suggests that temperature should decline over time according to the
following model.** $$\begin{aligned}
\mbox{temperature} & = & 60 +70e^{-\theta \;\mbox{elapsed.time}}
\end{aligned}$$ **You are to evaluate this model, i.e., make estimates
of** $\theta$ **using classical and Bayesian techniques.**

**2.1. Plot temperature against time (use the `scatter.smooth` function
to draw a nonparametric regression line through the points).**

```{r}
plot(elapsed.time, temperature, main="Temperature of the bore against elapsed time", pch=19, col="grey40")
```

**2.2. Fit the model in (i) using a classical approach that assumes that
observations have model errors are iid Normal(0,**$\sigma^2$).
$$\begin{aligned}
 \mbox{temperature} & \sim & \mbox{Normal} \left ( 60 +70e^{-\theta \; \mbox{elapsed.time}}, \sigma^2 \right )
\end{aligned}$$ Use the `nls` function in R. The format of `nls` in this
case:
`nl.1 <- nls(formula= temperature ~ 60 + 70*exp(-theta*elapsed.time), start=list(theta=initial.theta))`
**where `initial.theta` is an initial guess as to what** $\theta$ is.

**One way to get an estimate of** $\theta$ **is to "linearize" Newton's
law of cooling as follows:** $$\begin{aligned}
-\ln \left ( \frac{(\mbox{temperature}-60)}{70} \right )  & = &  \theta*\mbox{elapsed.time}
\end{aligned}$$ **and then fit the resulting linear model with the `lm`
function.**

**Use the estimated coefficient in `out` as the value of
`initial.theta`. After fitting the model, plot the fit and the
observations.**

```{r}
y <- - log((temperature - 60) / 70)
out <- lm(y ~ -1 + elapsed.time)
initial.theta <- out$coefficients["elapsed.time"]
nl.1 <- nls(formula=temperature ~ 60 + 70 * exp(- theta * elapsed.time), start=list(theta=initial.theta))
summary(nl.1)
```

```{r}
plot(temperature ~ elapsed.time,xlab="Time",ylab="Temperature", main="Friction Experiment Data")
lines(elapsed.time, fitted(nl.1), col="red")
# not good fit
```

**How does the fit look?**

**2.3. Instead of using 60 and 70 in Newton's law of cooling as known
values, refit the model estimating the coefficients (these coefficients
need to be assigned some parameter names such as beta0 and beta1, and
then included in the start parameter of the nls function, see ?nls for
more info).**

**Compare the estimated coefficients to the assumed values and plot the
fitted line over the top of the previous plot. Has the fit improved?**

```{r}
nl.2 <- nls(temperature ~ beta0 + beta1 * exp(- theta * elapsed.time), start=list(beta0=50, beta1=50, theta=initial.theta))
summary(nl.2)
plot(temperature ~ elapsed.time,xlab="Time",ylab="Temperature", main="Friction Experiment Data")
lines(elapsed.time, fitted(nl.2), col="red")
# better
```

**2.4. Use JAGS to fit two Bayesian nonlinear regression models: one
based on Newton's law of cooling, as in (ii), and another where all
three coefficients are estimated, as in (iii). Assume that temperatures
are normally distributed in the likelihood model.**

**In both cases use exponential distributions for the priors for**
$\theta$ **and then for** $\beta_0$ **and** $\beta_1$ **(to ensure that
the posterior distributions are positive valued). To pick the
exponential distribution hyperparameter, say** $\alpha$**, note that
if** $X \sim$**Exp**$(\alpha)$**,** $\mathbb{E}[X]$ **=** $1/\alpha$**.
Pick a large value for the hyperparameter for** $\theta$ **such that the
expected value of** $\theta$ **is less than 1. For the 2nd model (as in
(iii)), select hyperparameter values for** $\beta_0$ **and** $\beta_1$
**such that the expected values are 60 and 70, respectively. Note: in
JAGS, the exponential density is written `theta ~ dexp(a)` given
hyperparameter** $a$**.**

**Compare the posterior means for** $\theta$ **in both models with the
frequentist estimates.\
Likewise compare the posterior means for** $\beta_0$ **and** $\beta_1$
**for the second model.**

```{r}
model_string1 <- "
model {
  tau ~ dgamma(a, b)
  sigma = 1 / sqrt(tau)
  
  theta ~ dexp(alpha)
  
  for (i in 1: n) {
    mu[i] = 60 + 70 * exp(- theta * elapsed.time[i])
    temp[i] ~ dnorm(mu[i], tau)
  }
}
"

data <- list(elapsed.time=elapsed.time, temp=temperature, n=13, a=0.1, b=0.1, alpha=10)
model.j1 <- jags.model(textConnection(model_string1), data=data, n.chains=3)  # inits not specified
update(model.j1, n.iter=1000)
res.j1 <- coda.samples(model.j1, n.iter=10000, variable.names=c("theta"))
plot(res.j1)
summary(res.j1)
```

```{r}
model_string2 <- "
model {
  tau ~ dgamma(a, b)
  sigma = 1 / sqrt(tau)
  
  theta ~ dexp(alpha)
  
  b0 ~ dexp(alpha0)
  b1 ~ dexp(alpha1)
  
  for (i in 1: n) {
    mu[i] = b0 + b1 * exp(- theta * elapsed.time[i])
    temp[i] ~ dnorm(mu[i], tau)
  }
}
"

data <- list(elapsed.time=elapsed.time, temp=temperature, n=13, a=0.1, b=0.1, alpha=10, alpha0=1 / 60, alpha1=1 / 70)
model.j2 <- jags.model(textConnection(model_string2), data=data, n.chains=3)
update(model.j2, n.iter=10000)
res.j2 <- coda.samples(model.j2, n.iter=10000, variable.names=c("theta", "b0", "b1"))
par(mfrow=c(3, 2))
plot(res.j2)
summary(res.j2)
```

**3. Multiple Linear Regression\
Factors Affecting Extinction Times of 62 Land Bird Species, adapted from
Albert, 2009\
The data are taken from Ramsey and Schafer (1997), who took them from
Pimm et al. 1988, and are available in the `LearnBayes` package as the
object `birdextinct`. Land birds on 16 small islands had been observed
annually during breeding surveys over a period of several decades. Some
62 species went extinct at some point and the objective is to examine
the relationship between the years till extinction and three different
covariates: the initial average number of nesting pairs observed
(`nesting`), the physical size of the birds (an indicator variable
`size` with 1=small and 0=large), and migratory status (an indicator
variable `status` with 1=resident, 0=migratory).**

**To begin, do the following in `R`.**

```{r}
library(LearnBayes)

data(birdextinct)
n <- nrow(birdextinct)
extinct.time <- birdextinct$time
avg.no.nests <- birdextinct$nesting
size.ind <- birdextinct$size   # 0 = large, 1= small
mig.ind <- birdextinct$status # 0 = mig, 1=resident

size <- rep("Small", length(size.ind))
size[which(size.ind == 0)] <- "Large"
mig <- rep("Resident", length(mig.ind))
mig[which(mig.ind == 0)] <- "Mig"
```

**3.1. We start by doing some exploratory data analysis and data
transformation. Look at the histogram of `extinct.time`. It is strongly
right skewed (there are few species with times till extinction that are
long relative to most species). Therefore make the response variable the
natural log of exinct.time:**

```{r}
log.extinct <- log(extinct.time)
```

**3.2. Make 4 plots of** $y$**=`log.extinct`: histogram of** $y$ **,
scatterplot of** $y$ **against `avg.nests`, side-by-side boxplots of**
$y$ **for small and large birds, and side-by-side boxplots of resident
and migratory birds. Hint: To make side-by-side boxplots use the `split`
function; e.g.,`boxplot(split(log.extinct,size),main='vs Size')`**

```{r}
par(mfrow=c(2, 2))
par(mar=c(2, 2, 2, 2))
hist(log.extinct, main="Histogram")
scatter.smooth(avg.no.nests, log.extinct, main="Scatterplot")
boxplot(split(log.extinct, size), main="Boxplot - Size")
boxplot(split(log.extinct, mig), main="Boxplot - Mig")
```

**3.3. How would you describe the relationships between the 3 covariates
and time till extinction?**

**3.4. Fit a classical multiple linear regression of the `log.extinct`
on the three covariates**

```{r}
extinct.mlr <- lm(log.extinct ~ avg.no.nests + size + mig)
```

**3.5. Examine the estimated coefficients. How do they compare to your
conclusions from the EDA (Exploratory Data Analysis)?**

```{r}
summary(extinct.mlr)
```

**In questions 3.6-3.11, we use JAGS to fit a Bayesian multiple
regression analysis.**

**3.6. Implement the Bayesian version of the above multivariate
regression model in JAGS. Centre the *avg.no.nests* covariate. Use 3
sets of initial values for the parameters.**

```{r}
avg.no.nests.cent <- avg.no.nests - mean(avg.no.nests)
model_string1 <- "
model {
  tau ~ dgamma(a, b)
  sigma = 1 / sqrt(tau)
  
  for (i in 1: 4) {
    beta[i] ~ dnorm(mu0, tau0)
  }
  
  for (i in 1: n) {
    mu[i] = beta[1] + beta[2] * avg.no.nests[i] + beta[3] * size[i] + beta[4] * mig[i]
    log.extinct[i] ~ dnorm(mu[i], tau)
  }
}
"
data <- list(log.extinct=log.extinct, avg.no.nests=avg.no.nests.cent, size=size.ind, mig=mig.ind, n=n, a=0.01, b=0.01, mu0=0, tau0=1e-5)
my.inits <- list(list(tau=0.1, beta=c(0.1, 1, 10, 3)),
                 list(tau=0.1, beta=c(0.2, -5, 5, 1)),
                 list(tau=0.2, beta=c(0.2, 10, 20, -2)))
model.j1 <- jags.model(textConnection(model_string1), data=data, n.chains=3, inits=my.inits)
update(model.j1, n.iter=1000)
res.j1 <- coda.samples(model.j1, n.iter=10000, variable.names=c("sigma", "beta"))
```

**3.7. Plot the JAGS output to see the trace plots. (Type
`par(ask=TRUE)` in order to see all 5 plots. Then type `par(ask=FALSE)`
to turn the option off.)**

```{r}
par(mfrow=c(5, 2))
par(mar=c(2, 2, 2, 2))
plot(res.j1)
```

**3.8. Use the Gelman-Rubin diagnostics to check for convergence.**

```{r}
gelman.diag(res.j1)
gelman.plot(res.j1)
```

**3.9. Plot the autocorrelation functions.**

```{r}
par(mfrow=c(3, 2))
par(mar=c(1, 1, 1, 1))
acf(res.j1[[1]][, "beta[1]"], lag.max=30)
acf(res.j1[[1]][, "beta[2]"], lag.max=30)
acf(res.j1[[1]][, "beta[3]"], lag.max=30)
acf(res.j1[[1]][, "beta[4]"], lag.max=30)
acf(res.j1[[1]][, "sigma"], lag.max=30)
```

**3.10. Examine the effective sample sizes.**

```{r}
effectiveSize(res.j1)
```

**3.11. Calculate studentised residuals, draw a QQ-plot to check
normality, plot posterior mean fitted values, and carry out posterior
predictive checks for the minimum and maximum log.extinct times. (See
the mtcars example in the R code for Lecture 3 on the Learn site for
example code to do this.)**

```{r}
# number of chains doesn't matter
res.j1.mat <- as.matrix(res.j1)  # (30000, 5)
nsamp <- nrow(res.j1.mat)
sigma <- res.j1.mat[, "sigma"]
x <- cbind(rep(1, n), avg.no.nests.cent, size.ind, mig.ind)
H <- x %*% solve(t(x) %*% x) %*% t(x)

fittedvalues <- x %*% t(res.j1.mat[, -5])  # (62, 30000)
studentizedred <- matrix(0, nrow=n, ncol=nsamp)
for (i in 1: n) 
  for (j in 1: nsamp)
    studentizedred[i, j] <- (log.extinct[i] - fittedvalues[i, j]) / (sigma[j] * sqrt(1 - diag(H)[i]))  # (62, 30000)
studentizedredm <- apply(studentizedred, 1, mean)  # (62)
```

```{r}
qqnorm(studentisedresidm)
qqline(studentisedresidm)
```

```{r}
fittedvaluesm <- apply(fittedvalues, 1, mean)
plot(fittedvaluesm, studentisedresidm)
```

```{r}
yrep <- matrix(0, nrow=n, ncol=nsamp)
for (i in 1: nsamp) {
  yrep[, i] <- rnorm(n, mean=fittedvalues[, i], sd=sigma[i])
}
yrepmin <- apply(yrep, 2, min)
yrepmax <- apply(yrep, 2, max)
# note the dimension here! should be the max & min of the n data, not nsamp samples!
par(mfrow=c(2, 1))
hist(yrepmin, col="grey40")
abline(v=min(log.extinct), col="red")
hist(yrepmax, col="grey40")
abline(v=max(log.extinct), col="red")
```

**4. State space modeling for gray whale abundance**

**In this exercise, we continue work with our state space model from
Lecture 3.**

**As a recap, we have some observations of gray whale abundance (i.e.
population size) during some of the years in the period 1951-1997. It
turns out that it is easier to fit a model on the logarithm of the
population rather than the population directly. We are going to use the
following models from the true log-population** $x_t$ **(**$t$ **denotes
the year),**

$$x_t = b x_{t-1}+u+w_t; \quad w_t\sim N(0,\sigma^2).$$

**The observations are not assumed to be exact, but we allow them to
have an error, this is modelled as**

$$y_t =x_{t}+v_t, \quad v_t\sim N(0,\eta^2).$$

**This is a simple State-Space Model (SSM), also called Hidden Markov
Model.** $x_t$ **are the hidden states, and** $y_t$ **are the
observations.**

The following lines load the dataset and create the observation vector
$y$.

```{r}
library("MARSS")
#This loads to MARSS library
#If not available, please use install.packages("MARSS") first before loading it

# We load the graywhales dataset, and print out the values
data(graywhales)
print(graywhales)

n = 1997 - 1951 + 1
#We will have one y[i] for each year from 1951 until 1997
#So i=1 corresponds to 1951, i=2 corresponds to 1952, etc.

y=rep(NA, n)
for(i in 1: nrow(graywhales)) {
    y[graywhales[i, 1] - 1950] = log(graywhales[i, 2])
}
```

**4.1. Implement the model in JAGS using the following prior
distributions for the model parameters:**

-   **Gaussian prior** $N(\log(2500),1)$ for the initial state $x_0$,

-   **Uniform prior on the interval (0,1) for parameter** $b$,

-   **Exponential prior with parameter 1 for** $u$,

-   **Inverse Gamma prior with parameters (0.1,0.1) for** $\sigma^2$ and
    $\eta^2$.

**Hint: you can use the code provided with Lecture 3 for this.**

**Compile and run the JAGS simulations. Compute the effective samples
sizes for these 5 parameters. Choose the number of steps in the burn-in
period and the number of MCMC iterations in a way to ensure that the
effective sample sizes for the 4 parameters** $x_0$,$b$**,** $u$**,**
$\sigma^2$ **and** $\eta^2$ **are all above 1000. Once this is ensured,
compute summary statistics and plot the posterior densities of these
parameters.**

```{r}
model_string <- "
model {
  x[1] ~ dnorm(mu0, tau0)
  y[1] ~ dnorm(x[1], tauy)
  yrep[1] ~ dnorm(x[1], tauy)
  b ~ dunif(ab, bb)
  u ~ dexp(alpha)
  
  taux ~ dgamma(ax, bx)
  sigma2 = 1 / taux
  tauy ~ dgamma(ay, by)
  eta2 = 1 / tauy
  
  for (i in 2: n) {
    x[i] ~ dnorm(b * x[i - 1] + u, taux)
    y[i] ~ dnorm(x[i], tauy)
    yrep[i] ~ dnorm(x[i], tauy)
  }
}
"
library(imputeTS)
x.init=na_interpolation(y)

data <- list(y=y, n=n, ab=0, bb=1, mu0=log(2500), tau0=1, alpha=1, ax=0.1, bx=0.1, ay=0.1, by=0.1)

# init values can affect effective sample size!
my.inits <- list(list(b=0.8, u=0.5, taux=1, tauy=1, x=x.init+rnorm(n, sd=0.1)),
                   list(b=0.9, u=0.3, taux=2, tauy=2, x=x.init+rnorm(n, sd=0.1)),
                   list(b=0.7, u=1, taux=0.7, tauy=0.7, x=x.init+rnorm(n,sd=0.1)),
                   list(b=0.6, u=1.5, taux=0.5, tauy=0.5, x=x.init+rnorm(n,sd=0.1)),
                    list(b=0.5, u=2, taux=0.3, tauy=0.3, x=x.init+rnorm(n,sd=0.1)))

model.j <- jags.model(textConnection(model_string), data=data, n.chains=5, inits=my.inits)
update(model.j, n.iter=30000)
res.j <- coda.samples(model.j, n.iter=170000, variable.names=c("x[1]", "b", "u", "sigma2", "eta2"))
effectiveSize(res.j)
summary(res.j)
plot(res.j)
```

**4.2. Now we are going to perform prior sensitivity checks. Try a
different prior distribution for the 5 variables. For** $x_0$**, please
do not change the mean significantly from log(2500), but you can try a
different variance. When formulating priors, you are encouraged to use
your understanding of the model parameters and biological facts (see
e.g. <https://en.wikipedia.org/wiki/Gray_whale>). Run the simulations
again, and compare the summary statistics and posterior density plots of
these 5 parameters.**

```{r}
data.m <- list(y=y, n=n, ab=0.5, bb=1, mu0=log(2500), tau0=2, alpha=6, ax=1, bx=1 / 20, ay=1, by=1 / 20)
model.j.m <- jags.model(textConnection(model_string), data=data.m, n.chains=5, inits=my.inits)
update(model.j.m, n.iter=30000)
res.j.m <- coda.samples(model.j.m, n.iter=170000, variable.names=c("x[1]", "b", "u", "sigma2", "eta2"))
effectiveSize(res.j.m)
summary(res.j.m)
plot(res.j.m)
```

**4.3.** **Update the model to compute the posterior distribution of the
log population sizes (**$x_t$**) every year between 1951-2050. Plot the
evolution of the posterior mean of the log population sizes from
1951-2050 on a single plot, along with two other curves that correspond
to the [2.5%, 97.5%] credible interval of the log population size
(**$x_t$**) according to the posterior distribution at each year.
Finally, estimate the posterior probability that the population of gray
whales becomes smaller than 100 at any year from 1951 until the end of
2050 (i.e.,** $p(\min_{t\in \{0,1\ldots,99\}} x_t<=\log(100)|y)$**).**

```{r}
n3 <- 100
y3 <- rep(NA, n3)
y3[1: n] <- y

require(imputeTS)
x.init3 <- na_interpolation(y3)

data.m <- list(y=y3, n=n3, ab=0.5, bb=1, mu0=log(2500), tau0=2, alpha=6, ax=1, bx=1 / 20, ay=1, by=1 / 20)

my.inits <- list(list(b=0.8, u=0.5, taux=1, tauy=1, x=x.init3 + rnorm(n3, sd=0.1)),
                     list(b=0.9, u=0.3, taux=2, tauy=2, x=x.init3 + rnorm(n3, sd=0.1)),
                     list(b=0.7, u=1, taux=0.7, tauy=0.7, x=x.init3 + rnorm(n3, sd=0.1)),
                     list(b=0.6, u=1.5, taux=0.5, tauy=0.5, x=x.init3 + rnorm(n3, sd=0.1)),
                     list(b=0.5, u=2, taux=0.3, tauy=0.3, x=x.init3 + rnorm(n3, sd=0.1)))

model3 <- jags.model(textConnection(model_string), data=data.m, n.chains=5, inits=my.inits)

update(model3, 30000)

# Running the model
res.model3 <- coda.samples(model3,variable.names=
           c("b", "u", "sigma2", "eta2", "x"), n.iter=170000)
```

```{r}
res.model3.mat <- as.matrix(res.model3)
nsamp <- nrow(res.model3.mat)
x.samples <- res.model3.mat[, 5: 104]
xm <- apply(x.samples, 2, mean)  # posterior mean for log value each year
x.quant <- apply(x.samples, 2, function(x) quantile(x, prob=c(0.025, 0.975)))
plot(1951: 2050, xm)
lines(1951: 2050, x.quant[1, ])
lines(1951: 2050, x.quant[2, ])

sum(apply(x.samples, 1, min) <= log(100))
```

**4.4. We are going to perform posterior predictive checks to evaluate
the fit of this model on the data. First, create replicate observations
from the posterior predictive using JAGS. The number of replicate
observations should be at least 1000. Compute the minimum, maximum,
median, kurtosis and skewness for the replicated observations. Plot the
histograms for these quantities together with a line that shows the
value of the function considered on the actual dataset (see the R code
for Lecture 3 for an example). Perform these checks both for the
original prior from Question 5.1, and the prior you have proposed in
Question 5.2. Discuss the results.**

```{r}
y.not.na.ind <- which(!is.na(y))
y.not.na <- y[!is.na(y)]

res.yrep <- coda.samples(model3, n.iter=100000, variable.names=c("yrep"))  # (nsamp, n)

res.yrep <- as.matrix(res.yrep)[, y.not.na.ind]

yrep.min <- apply(res.yrep, 1, min)
yrep.max <- apply(res.yrep, 1, max)
yrep.med <- apply(res.yrep, 1, median)
require(fBasics)
yrep.kurt <- apply(res.yrep, 1, kurtosis)
yrep.skew <- apply(res.yrep, 1, skewness)

par(mfrow=c(3,2))
hist(yrep.min,col="gray40",main="Predictive distribution for min")
abline(v=min(y.not.na),col="red",lwd=2)
hist(yrep.max,col="gray40",main="Predictive distribution for max")
abline(v=max(y.not.na),col="red",lwd=2)
hist(yrep.med,col="gray40",main="Predictive distribution for median")
abline(v=median(y.not.na),col="red",lwd=2)
hist(yrep.kurt,col="gray40",main="Predictive distribution for kurtosis")
abline(v=kurtosis(y.not.na),col="red",lwd=2)
hist(yrep.skew,col="gray40",main="Predictive distribution for skewness")
abline(v=skewness(y.not.na),col="red",lwd=2)
```
